---
title: "Exploratory Data Analysis of Periodontitis data"
author: "Martin Ruhle & Emiliano Jiménez"
date: "2024-08-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introducción

Vamos a buscar comprender como está compuesta la cohorte, así como las relaciones entre las variables.


```{r warning=FALSE, messAGE=FALSE}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(polycor)
library(GGally)
```

### Data Loading and Cleaning

We load the metadata file and perform basic cleaning step

```{r eval=FALSE}
# Read the data
metadata <- read.csv("metadata.csv")
```
```{r echo=FALSE}
metadata <- read.csv("C:/Users/User/Documents/Ecología_microbiana/Proyecto_final/GitHub_repository/Data/completo_full.csv")
```
```{r}
# FIltering metdata variables
metadata_fil <- metadata %>%
    select(Run, AGE, BOP_total, CAL_total, Cigarrettes_day, health_condition, Months_smoking, geo_loc_name_country, health_condition, RUN_ID, sex, Smoking, teeth, Periodontitis_extent_calc, Periodontitis_extent, Periodontitis_severity, )

# Basic Statistics and Structure
str(metadata_fil)
summary(metadata_fil)

# Checking for missing values
colSums(is.na(metadata_fil))
```

### Clean and Tidy the Metadata

```{r}
# Extract the mock samples
mock <- metadata_fil %>%
  filter(is.na(health_condition))

metadata_fil <- metadata_fil %>%
  filter(!is.na(health_condition))
```


```{r warning=FALSE}
# Replace empty strings with "Healthy" in the Periodontitis_extent column
metadata_fil$Periodontitis_extent[metadata_fil$Periodontitis_extent == ""] <- "Healthy"

metadata_fil <- metadata_fil %>%
  mutate(
    cigarrettes_group = case_when(
      Cigarrettes_day == 0 ~ "No smoker",
      Cigarrettes_day > 0 & Cigarrettes_day <= 10 ~ "1-10",
      Cigarrettes_day > 10 & Cigarrettes_day <= 20 ~ "11-20",
      Cigarrettes_day > 20 ~ "20+",
      TRUE ~ NA_character_
    ),
    Months_smoking_group = case_when(
      Months_smoking == 0 ~ "No smoker",
      Months_smoking > 0 & Months_smoking <= 120 ~ "1-120",
      Months_smoking > 120 & Months_smoking <= 240 ~ "121-240",
      Months_smoking > 240 & Months_smoking <= 360 ~ "241-360",
      Months_smoking > 360 ~ "360+"
      ),
    teeth_group = case_when(
      teeth < 20 ~ "10-19",
      teeth >= 20 & teeth < 25 ~ "20-24",
      teeth >= 25 & teeth < 30 ~ "25-29",
      teeth >= 30 ~ "30+"
      ),
    BOP_group = case_when(
      BOP_total <= 10 ~ "0-10",
      BOP_total > 10 & BOP_total <= 30 ~ "11-30",
      BOP_total > 30 & BOP_total <= 60 ~ "31-60",
      BOP_total > 60 ~ "61+"
    ),
    CAL_group = case_when(
      CAL_total <= 2.5 ~ "1.3-2.5",
      CAL_total > 2.5 & CAL_total <= 4 ~ "2.51-4",
      CAL_total > 4 & CAL_total <= 6 ~ "4.01-6",
      CAL_total > 6 ~ "6.01+"
    ))
```


### Data Analysis
A generalized analysis is first preformed, visualizing and calculating correlations between the main variables.

```{r}
ggpairs(metadata_fil, columns = c(2:8, 10:13), ggplot2::aes(colour=factor(Periodontitis_extent, levels = c("Generalized", "Localized", "Healthy")), alpha = 0.75))
```

## Demographic Analysis
## AGE Distribution
We analyze the distribution of participants' AGEs and the relationship between AGE and other variables. 

```{r warning=FALSE}
# AGE Participant Distribution
ggplot(metadata_fil, aes(x = AGE)) + 
  geom_histogram(stat = "count", fill = "#5ea758", color = "grey") + 
  xlab("AGE") + 
  ylab("Count") + 
  ggtitle("AGE Participant Distribution")

# Define the AGE groups
AGE_groups <- metadata_fil %>%
  filter(!is.na(AGE)) %>%
  mutate(AGE_group = case_when(
    AGE < 20 ~ "<20",
    AGE >= 20 & AGE < 30 ~ "20-29",
    AGE >= 30 & AGE < 40 ~ "30-39",
    AGE >= 40 ~ "40+"
  ))

# Calculate the percentAGE for each AGE group
AGE_group_percentAGEs <- AGE_groups %>%
  count(AGE_group) %>%
  mutate(percentAGE = n / sum(n) * 100)

# Display the results
AGE_group_percentAGEs

```

### AGE and Preterm Correlation Analysis


```{r warning=FALSE}
# Convert logical columns to numeric
metadata_fil$periodontitis <- as.numeric(metadata_fil$health_condition)

# Calculate the correlations
AGE_periodontitis_cor <- polyserial(metadata_fil$AGE, factor(metadata_fil$health_condition))

# Calculate the p-values
smocking_bop_test <- cor.test(metadata_fil$Months_smoking, metadata_fil$BOP_total, method = "pearson")

smocking_bop_test_generalized <- cor.test(metadata_fil[metadata_fil$Periodontitis_extent == "Generalized",]$Months_smoking, metadata_fil[metadata_fil$Periodontitis_extent == "Generalized",]$BOP_total, method = "pearson")

# Create a formatted matrix with correlations and significance levels
pvalue_polyserial <- paste0(round(AGE_periodontitis_cor, 2), " (p=", format(AGE_periodontitis_test$p.value, digits = 2), ")")
pvalue_person <-
formatted_matrix[1, 2] <- paste0(round(AGE_periodontitis_cor, 2), " (p=", format(AGE_periodontitis_test$p.value, digits = 2), ")")

ggplot(metadata_fil, aes(x=Months_smoking, y=BOP_total, color=Periodontitis_extent)) +
  geom_point() +
  geom_smooth(method = "lm", )

ggplot(metadata_fil, aes(x=Months_smoking, y=CAL_total, color=Periodontitis_extent)) +
  geom_point() +
  geom_smooth(method = "lm")

ggplot(metadata_fil, aes(x=Months_smoking, y=Periodontitis_extent_calc, color=Periodontitis_extent)) +
  geom_point() +
  geom_smooth(method = "lm")

# Overlaid AGE Participant Distribution by was_term
ggplot(metadata_fil[!is.na(metadata_fil$AGE),], aes(x = AGE_factor, fill = as.factor(was_term))) + 
  geom_histogram(stat = "count", position = "identity", alpha = 0.5, color = "black") + 
  xlab("AGE") + 
  ylab("Count") + 
  ggtitle("AGE Participant Distribution by Term Status") +
  scale_fill_manual(name = "Term Status", values = c("TRUE" = "blue", "FALSE" = "red"), labels = c("TRUE" = "Term", "FALSE" = "Preterm")) + 
  annotate("text", x = Inf, y = Inf, label = paste("point-biserial corr:", format(formatted_matrix[2,1], digits = 2)), hjust = 1.1, vjust = 2, size = 3)


# Overlaid AGE Participant Distribution by was_term
ggplot(metadata_fil[!is.na(metadata_fil$AGE),], aes(x = AGE_factor, fill = as.factor(was_early_preterm))) + 
  geom_histogram(stat = "count", position = "identity", alpha = 0.5, color = "black") + 
  xlab("AGE") + 
  ylab("Count") + 
  ggtitle("AGE Participant Distribution by Early Preterm Status") +
  scale_fill_manual(name = "Early preterm Status", values = c("TRUE" = "red", "FALSE" = "blue"), labels = c("TRUE" = "Early preterm", "FALSE" = "Not early preterm")) + 
  annotate("text", x = Inf, y = Inf, label = paste("point-biserial corr:", format(formatted_matrix[3,1], digits = 2)), hjust = 1.1, vjust = 2, size = 3)
```

Like seen in bibliography the maternal AGE is correlated with the preterm risk (Fuchs 2018).

## Pregnancy and Delivery Details
### Term Status

We analyze the distribution of term, preterm, and early preterm statuses among participants.

```{r}
# Create the table
term_status <- table(participant_data$was_preterm, participant_data$was_early_preterm)

# Add descriptive names to the dimensions
dimnames(term_status) <- list(
  Preterm_Status = c("Full-Term", "Preterm"),
  Early_Preterm_Status = c("Not Early Preterm", "Early Preterm")
)

# Display the table with descriptive names
print(term_status)

# Calculate percentAGEs
total_births <- sum(term_status)
preterm_births <- sum(term_status["Preterm",])
early_preterm_births <- sum(term_status[, "Early Preterm"])

preterm_percentAGE <- (preterm_births / total_births) * 100
early_preterm_percentAGE <- (early_preterm_births / total_births) * 100

# Print the percentAGEs
cat(sprintf("PercentAGE of Preterm Births: %.2f%%\n", preterm_percentAGE))
cat(sprintf("PercentAGE of Early Preterm Births: %.2f%%\n", early_preterm_percentAGE))
```

These values are different from the values shown in the "2023 March of Dimes report card for United States" (<https://www.marchofdimes.org/peristats/reports/united-states/report-card>.), where the averAGE precentAGE was 10,4%.
The criteria of data collection probably searched to have a more balanced dataset to train the predictive model (Qiong Wei 2013) 

### Delivery Week Distribution

We analyze the distribution of delivery weeks among participants and how it relates to term status.

```{r}
# Plotting Delivery Week Distribution
ggplot(participant_data, aes(x = delivery_wk, fill = as.factor(was_term))) + 
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.5, color = "black") + 
  xlab("Delivery Week") + 
  ylab("Count") + 
  ggtitle("Distribution of Delivery Week") +
  scale_fill_manual(name = "Term Status", values = c("TRUE" = "blue", "FALSE" = "red"), labels = c("TRUE" = "Term", "FALSE" = "Preterm"))

```

```{r}
# Remove rows with NA values in AGE or delivery_wk
correlation_data <- participant_data %>% filter(!is.na(AGE), !is.na(delivery_wk))

# Check distribution of AGE and delivery_wk
hist(correlation_data$AGE, main="AGE Distribution", xlab="AGE")
hist(correlation_data$delivery_wk, main="Delivery Week Distribution", xlab="Delivery Week")

# Conduct normality tests (Shapiro-Wilk test)
shapiro_AGE <- shapiro.test(correlation_data$AGE)
shapiro_delivery_wk <- shapiro.test(correlation_data$delivery_wk)

# Print normality test results
shapiro_AGE
shapiro_delivery_wk

# Choose correlation test based on normality
if (shapiro_AGE$p.value > 0.05 & shapiro_delivery_wk$p.value > 0.05) {
  # Use Pearson correlation if both are normally distributed
  correlation_test <- cor.test(correlation_data$AGE, correlation_data$delivery_wk, method = "pearson")
} else {
  # Use Spearman correlation if not normally distributed
  correlation_test <- cor.test(correlation_data$AGE, correlation_data$delivery_wk, method = "spearman")
}

# Print the correlation test results
correlation_test
```

The Spearman's correlation test p-value result indicates the statistical of the correlation between AGE and delivery week. 
A value of -0.1035 in the Correlation Coefficient (rho) suggests a weak, negative correlation. This means that as AGE increases, delivery_wk tends to slightly decrease.

### Collection Week Distribution
We analyze the distribution of samples collection weeks, both the earliest and latest collection weeks, among participants.

```{r}
# Plotting sample collection weeks
ggplot(metadata, aes(x = collect_wk)) + 
  geom_histogram(binwidth = 1, fill = "#5ea758", color = "grey") + 
  xlab("Collection Week") + 
  ylab("Count") + 
  ggtitle("Distribution of Sample Collection Week")
```

There is a big peak of samples taken in the week 18, most of them come from only one study the "I".

```{r}
# Table indicating the sample week collection of the I project 
table(metadata[metadata$project=="I",]$project, metadata[metadata$project=="I",]$collect_wk)

# Table of participants samples per project
table(participant_data$n_samples, participant_data$project)

```

The "I" project has the peculiarity of having more than one sample per participant per week. This could be useful for analyzing the method consistency.

```{r}
# Distribution of Number of Samples per Participant
ggplot(participant_data, aes(x = n_samples)) + 
  geom_histogram(binwidth = 1, fill = "#5ea758", color = "grey") + 
  xlab("Number of Samples per Participant") + 
  ylab("Count") + 
  ggtitle("Distribution of Number of Samples per Participant")
```
There are 889 participants that have more than one sample that represents the 70%. This could be used to make longitudinal analysis.

To disgregate the distribution of the earliest, latest and single samples we add an extra column.
```{r warning=FALSE}
# Categorize collection weeks
participant_data <- participant_data %>%
  mutate(
    collect_wk_category = case_when(
      collect_wk_earliest == collect_wk_latest ~ "single",
      TRUE ~ "multiple"
    )
  )

# Separate data for multiple and single categories
multiple_data <- participant_data %>% filter(collect_wk_category == "multiple")
single_data <- participant_data %>% filter(collect_wk_category == "single")

# Plotting Collection Week Distribution (Earliest and Latest)
ggplot() + 
  geom_histogram(data = multiple_data, aes(x = collect_wk_earliest, fill = "Earliest"), binwidth = 1, color = "grey", alpha = 0.5, position = "identity") + 
  geom_histogram(data = multiple_data, aes(x = collect_wk_latest, fill = "Latest"), binwidth = 1, color = "grey", alpha = 0.5, position = "identity") + 
  geom_histogram(data = single_data, aes(x = collect_wk_earliest, fill = "single"), binwidth = 1, color = "grey", alpha = 0.5, position = "identity") + 
  xlab("Collection Week") + 
  ylab("Count") + 
  ggtitle("Collection Week") +
  scale_fill_manual(name = "Collection Week Category", values = c("Earliest" = "blue", "Latest" = "red", "single" = "green"), 
                    labels = c("Earliest", "Latest", "Single")) +
  theme(legend.position = "right")
```


### Race/Ethnicity Distribution

We analyze the distribution of NIH racial categories among the participants and perform a chi-squared test to examine the association between NIH racial categories and term status.

```{r}
# Race/Ethnicity Distribution
ethnicity_distribution <- table(participant_data$nih_racial_category)
ethnicity_distribution
```


```{r warning=FALSE}
# Realizar el test chi-cuadrado
chi_sq_test <- chisq.test(table(participant_data$nih_racial_category, participant_data$was_term))

# Calcular residuos ajustados
residuals_table <- as.data.frame(as.table(chi_sq_test$stdres))
colnames(residuals_table) <- c("nih_racial_category", "was_term", "std_residual")

# Añadir columna de significancia
residuals_table <- residuals_table %>%
  mutate(significance = case_when(
    abs(std_residual) > 3.29 ~ "***",  # p < 0.001
    abs(std_residual) > 2.68 ~ "**",   # p < 0.01
    abs(std_residual) > 1.96 ~ "*",    # p < 0.05
    TRUE ~ ""
  ))

# Preparar los datos para las etiquetas de las barras
data_summary <- participant_data %>%
  group_by(nih_racial_category, was_term) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(nih_racial_category) %>%
  mutate(total_count = sum(count),
         percentAGE = count / total_count * 100) %>%
  filter(was_term == FALSE)  # Mantener solo los nacimientos prematuros

# Ordenar categorías por porcentaje de nacimientos prematuros
ordered_categories <- data_summary %>%
  arrange(percentAGE) %>%
  pull(nih_racial_category) %>%
  unique()

participant_data$nih_racial_category <- factor(participant_data$nih_racial_category, levels = ordered_categories)

# Combinar residuos con los datos para las etiquetas
data_summary <- data_summary %>%
  left_join(residuals_table %>% filter(was_term == FALSE), by = "nih_racial_category")

# Generar el gráfico con ggplot2
ggplot(participant_data, aes(x = nih_racial_category, fill = was_term)) + 
  geom_bar(position = "stack") + 
  geom_text(data = data_summary, 
            aes(x = nih_racial_category, 
                y = total_count, 
                label = paste0(sprintf("%.1f%%", percentAGE), " ", significance)), 
            inherit.aes = FALSE, # Evitar heredar estéticas globales
            vjust = -0.5) +
  xlab("NIH Racial Category") + 
  ylab("Count") + 
  ggtitle("Distribution of NIH Racial Category with Preterm Proportions") +
  annotate("text", x = Inf, y = Inf, 
           label = paste("chi_sq p-value:", format(chi_sq_test$p.value, digits = 2)), 
           hjust = 1.1, vjust = 2, size = 3) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))


ggplot(participant_data, aes(x = nih_racial_category, fill = was_term)) + 
  geom_bar(position = "stack") + 
  # Agregar porcentajes debajo
  geom_text(data = data_summary, 
            aes(x = nih_racial_category, 
                y = total_count, 
                label = sprintf("%.1f%%", percentAGE)), 
            inherit.aes = FALSE, 
            vjust = -0.5) +
  # Agregar asteriscos encima
  geom_text(data = data_summary, 
            aes(x = nih_racial_category, 
                y = total_count + 20, # Ajusta esta distancia según el tamaño del gráfico
                label = significance), 
            inherit.aes = FALSE, 
            vjust = -0.5) +
  xlab("NIH Racial Category") + 
  ylab("Count") + 
  ggtitle("Distribution of NIH Racial Category with Preterm Proportions") +
  annotate("text", x = Inf, y = Inf, 
           label = paste("chi_sq p-value:", format(chi_sq_test$p.value, digits = 2)), 
           hjust = 1.1, vjust = 2, size = 3) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

```

We can see that there is a significant difference between the participants ethnic groups, in number and proportions of preterm births.

Of all live births in the United States during 2020-2022 (averAGE), 24.3% were Hispanic, 51.5% were White, 14.4% were Black, 0.7% were American Indian/Alaska Native and 6.4% were Asian/Pacific Islander (<https://www.marchofdimes.org/peristats/data?reg=99&top=2&stop=2&lev=1&slev=1&obj=1>)

On the other hand the ethnic percetAGEs of preterm birth are really different shown in the "2023 March of Dimes report card for United States" (<https://www.marchofdimes.org/peristats/reports/united-states/report-card>). Not just the percentAGEs, but also the order of the ethnic categories.

* Ethnic preterm precentAGEs:
  + Asian -- 9.0%
  + White -- 9.4%
  + Hispanic -- 10.1%
  + American Indian/Alaska Native -- 12.2%
  + Pacific Islanders -- 12.2%
  + Black -- 14.6%
  
  
References

Fuchs 2018 - <https://doi.org/10.1371/journal.pone.0191002>
Qiong Wei 2013 - <https://doi.org/10.1371/journal.pone.0067863>