---
title: "PCA Analysis of Periodontitis Relative Abundance Data"
author: "Martin Ruhle"
date: "2024-11-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This analysis aims to perform a Principal Component Analysis (PCA) on the relative abundance data of microbiome samples. The PCA will reveal patterns of variation in the data, which are visualized using various categorical variables.


### Load Libraries
We begin by loading the necessary libraries. `ggplot2` is used for visualization, `dplyr` for data manipulation, and `FactoMineR` for multivariate analysis.
```{r message=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(FactoMineR)
```

### Load the Data

```{r}
rel_abun_metadata <- read.csv("C:/Users/User/Documents/Ecología_microbiana/Proyecto_final/GitHub_repository/Data/genus_rel_abund_meta.csv")
```

```{r}
rel_abun_metadata <- rel_abun_metadata %>%
  mutate(
    cigarrettes_group = case_when(
      cigarrettes_group == 0 ~ "0",
      cigarrettes_group > 0 & cigarrettes_group <= 10 ~ "<=10",
      cigarrettes_group > 10 & cigarrettes_group <= 20 ~ "<=20",
      cigarrettes_group > 20 ~ "mas de 20",
      TRUE ~ NA_character_
    ))
```


### PCA Data Preparation
To prepare the data for PCA:

1. We exclude non-numeric columns.
2. We remove columns with zero variance, as they do not contribute to PCA.
3. We normalize the remaining numeric columns using z-score normalization.

```{r}
# Exclude non-numeric columns for PCA
pca_data <- rel_abun_metadata %>%
  dplyr::select(-c(1:20))

# Remove columns with zero variance
pca_data <- pca_data[, sapply(pca_data, function(col) var(col, na.rm = TRUE) > 0)]

# Normalize the data (z-score normalization)
pca_data_normalized <- scale(pca_data)
```

### Perform PCA
We perform two PCA analyses:

1. Unnormalized PCA: For the raw data.
2. Normalized PCA: Using z-score normalized data to account for scale differences.

```{r}
# Perform PCA on unnormalized data
pca_result <- prcomp(pca_data, scale. = FALSE)

# Perform PCA on normalized data
pca_result_norm <- prcomp(pca_data_normalized, scale. = TRUE)
```

### PCA Scores
We extract the PCA scores for the first two principal components and add relevant metadata columns for coloring in the visualizations.

```{r}
# Get the PCA scores for the first two principal components
pca_scores <- as.data.frame(pca_result$x[, 1:2])
pca_scores_norm <- as.data.frame(pca_result_norm$x[, 1:2])

# Add metadata columns for plotting
pca_scores <- pca_scores %>%
  mutate(
    Age = rel_abun_metadata$AGE,
    BOP_total = rel_abun_metadata$BOP_total,
    CAL_total = rel_abun_metadata$CAL_total,
    Cigarrettes_day = rel_abun_metadata$Cigarrettes_day,
    health_condition = rel_abun_metadata$health_condition,
    Months_smoking = rel_abun_metadata$Months_smoking,
    Country = rel_abun_metadata$geo_loc_name_country,
    Smoking = rel_abun_metadata$Smoking,
    teeth = rel_abun_metadata$teeth,
    Periodontitis_extent = rel_abun_metadata$Periodontitis_extent,
    cigarrettes_group = rel_abun_metadata$cigarrettes_group,
    Months_smoking_group = rel_abun_metadata$Months_smoking_group,
    teeth_group = rel_abun_metadata$teeth_group,
    BOP_group = rel_abun_metadata$BOP_group,
    CAL_group = rel_abun_metadata$CAL_group
  )

pca_scores_norm <- pca_scores_norm %>%
  mutate(
    project = rel_abun_metadata$project,
    specimen = rel_abun_metadata$specimen,
    age = rel_abun_metadata$age,
    was_preterm = rel_abun_metadata$was_preterm,
    was_early_preterm = rel_abun_metadata$was_early_preterm,
    NIH.Racial.Category = rel_abun_metadata$NIH.Racial.Category,
    age_group = rel_abun_metadata$age_group
  )
```

### PCA Plotting Function
To make the visualizations, we define a function `plot_pca()` that takes in the PCA data, a variable for coloring, and a title. This allows us to plot PCA results by various metadata categories.

```{r}
# Function to plot PCA with specified color variable
plot_pca <- function(data, color_var, title) {
  # Check if the variable is numeric or categorical
  if (is.numeric(data[[color_var]])) {
    # Continuous variable
    color_scale <- scale_color_gradient2(
      low = "blue",
      mid = "green",
      high = "red",
      midpoint = median(data[[color_var]], na.rm = TRUE),
      na.value = "grey50"
    )
  } else {
    # Categorical variable
    color_scale <- scale_color_discrete() # Uses the default ggplot2 palette
  }
  
  ggplot(data, aes(x = PC1, y = PC2, color = data[[color_var]])) +
    geom_point(alpha = 0.85) +
    labs(
      title = title,
      x = paste0("PC1 (", pc1_var, "%)"),
      y = paste0("PC2 (", pc2_var, "%)"),
      color = color_var
    ) +
    color_scale +
    theme_minimal(base_size = 14) +
    theme(
      panel.grid.major = element_line(color = "grey80"),
      panel.grid.minor = element_line(color = "grey90"),
      panel.background = element_rect(fill = "grey95"),
      plot.background = element_rect(fill = "grey95", color = NA),
      legend.background = element_rect(fill = "grey90", color = NA)
    )
}
```

### Visualizations
We use `plot_pca()` to create PCA plots, each colored by a different variable of interest: project, age, was_preterm, and NIH.Racial.Category.

```{r warning=FALSE}
# Plot PCA colored by each variable of interest
pca_plots <- list(
  age = plot_pca(pca_scores, "Age", "PCA of Relative Abundance Data (by Age)"),
  bop_total = plot_pca(pca_scores, "BOP_total", "PCA of Relative Abundance Data (by BOP Total)"),
  cal_total = plot_pca(pca_scores, "CAL_total", "PCA of Relative Abundance Data (by CAL Total)"),
  cigarrettes_day = plot_pca(pca_scores, "Cigarrettes_day", "PCA of Relative Abundance Data (by Cigarettes/Day)"),
  health_condition = plot_pca(pca_scores, "health_condition", "PCA of Relative Abundance Data (by Health Condition)"),
  months_smoking = plot_pca(pca_scores, "Months_smoking", "PCA of Relative Abundance Data (by Months Smoking)"),
  country = plot_pca(pca_scores, "Country", "PCA of Relative Abundance Data (by Country)"),
  smoking = plot_pca(pca_scores, "Smoking", "PCA of Relative Abundance Data (by Smoking Status)"),
  teeth = plot_pca(pca_scores, "teeth", "PCA of Relative Abundance Data (by Teeth Count)"),
  periodontitis_extent = plot_pca(pca_scores, "Periodontitis_extent", "PCA of Relative Abundance Data (by Periodontitis Extent)"),
  cigarrettes_group = plot_pca(pca_scores, "cigarrettes_group", "PCA of Relative Abundance Data (by Cigarettes Group)"),
  months_smoking_group = plot_pca(pca_scores, "Months_smoking_group", "PCA of Relative Abundance Data (by Months Smoking Group)"),
  teeth_group = plot_pca(pca_scores, "teeth_group", "PCA of Relative Abundance Data (by Teeth Group)"),
  bop_group = plot_pca(pca_scores, "BOP_group", "PCA of Relative Abundance Data (by BOP Group)"),
  cal_group = plot_pca(pca_scores, "CAL_group", "PCA of Relative Abundance Data (by CAL Group)")
)

# Display the plots
for (plot_name in names(pca_plots)) {
  print(pca_plots[[plot_name]])
}

```


### Disagregation and detailed PCA

Taking into account that "H" an "I" projects doesn´t have numeric ages we can plot separately.

```{r warning=FALSE}
# Filter out projects "H" and "I" before plotting
pca_scores_age_filtered <- pca_scores %>%
  filter(!project %in% c("H", "I"))

# Generate the PCA plot with age as the color variable, excluding projects "H" and "I"
pca_plot_age_filtered <- plot_pca(pca_scores_age_filtered, "age", "PCA of Relative Abundance Data (by Age, excluding Projects H and I)")

pca_plot_age_H <- plot_pca(pca_scores[!is.na(pca_scores$age_group),], "age_group", "PCA of Relative Abundance Data (by Age, excluding Projects H and I)")

print(pca_plot_age_filtered)
print(pca_plot_age_H)
```


### Species Analysis

```{r}
# Transform the dataset to long format, specifying columns 8 onward for species
rel_abun_long <- rel_abun_metadata %>%
  pivot_longer(cols = 8:ncol(rel_abun_metadata), names_to = "species", values_to = "abundance")

# Define abundance categories with a specific order
rel_abun_long <- rel_abun_long %>%
  mutate(abundance_category = case_when(
    abundance == 1 ~ "1",
    abundance > 0.99 ~ ">0.99",
    abundance > 0.95 ~ ">0.95",
    abundance > 0.9 ~ ">0.9",
    abundance > 0.75 ~ ">0.75",
    abundance > 0.5 ~ ">0.5",
    TRUE ~ "<=0.5"
  ),
  # Set the order for abundance categories
  abundance_category = factor(abundance_category, levels = c("1", ">0.99", ">0.95", ">0.9", ">0.75", ">0.5", "<=0.5"))
  )

# Calculate counts and cumulative percentages
cumulative_percentages <- rel_abun_long %>%
  filter(abundance_category != "<=0.5") %>%  # Exclude "<=0.5" if not needed
  group_by(abundance_category) %>%
  summarize(count = n()) %>%
  arrange(abundance_category) %>%  # This will now follow the specified factor levels order
  mutate(cumulative_count = cumsum(count)) %>%  # Cumulative sum of counts
  mutate(percentage = (cumulative_count / 3578) * 100)

# Reorder abundance_category to be in descending order
cumulative_percentages <- cumulative_percentages %>%
  mutate(abundance_category = factor(abundance_category, levels = rev(levels(abundance_category))))

# Plot the cumulative percentages with "1" as a distinct category
ggplot(cumulative_percentages, aes(x = abundance_category, y = percentage, fill = abundance_category)) +
  geom_bar(stat = "identity") +
  labs(title = "Cumulative Percentage of Samples by Relative Abundance Category",
       x = "Relative Abundance Category",
       y = "Percentage of Samples") +
  #scale_fill_brewer(palette = "viridis") +
  theme_minimal() +
  theme(legend.position = "none")
```


```{r}
# Plot the Abundance Taxa Distribution for Species and Genus
ggplot() +
  geom_density(data = rel_abun_long[rel_abun_long$abundance > 0.5, ], 
               aes(x = abundance, color = "Species")) +
  #geom_density(data = rel_abun_long_gen[rel_abun_long_gen$abundance > 0.5, ], 
  #             aes(x = abundance, color = "Genus")) +
  scale_color_manual(name = "Taxa level", values = c("Species" = "lightgreen"
                                                     #, "Genus" = "red") # If genus is needed
                     )) +
  labs(title = "Abundance Taxa Distribution",
       x = "Abundance", 
       y = "Density") +
  theme_minimal()
```

Now we will analyze the most dominant bacteria species and compare their preterm proportion with the total samples proportion using the Fisher's exact test.

```{r}
# Create a table with counts by species and was_preterm for abundance > 0.95
species_table <- table(
  rel_abun_long[rel_abun_long$abundance > 0.95, ]$species, 
  rel_abun_long[rel_abun_long$abundance > 0.95, ]$was_preterm
)

# Convert the table to a data frame for easier manipulation
species_df <- as.data.frame(species_table)

# Rename columns to match "TRUE" and "FALSE"
colnames(species_df) <- c("Species", "was_preterm", "Count")

# Spread the "was_preterm" column to create separate "TRUE" and "FALSE" columns
species_counts <- species_df %>%
  tidyr::pivot_wider(names_from = was_preterm, values_from = Count, values_fill = 0) %>%
  rename(`True` = `True`, `False` = `False`)  %>%
  mutate(Total = `True` + `False`) %>%  # Calculate total counts
  arrange(desc(Total)) %>%              # Sort by total counts in descending order
  mutate(Preterm_perc = (True/Total*100)) %>% # Calculate the precentage
  select(-Total)                      # Remove the total column if not needed in the output

# Overall counts from metadata
overall_counts <- table(metadata$was_preterm)
total_false <- overall_counts["False"]
total_true <- overall_counts["True"]

# Perform Fisher's exact test for each species
species_p_values <- species_counts %>%
  rowwise() %>%
  mutate(
    p_value = fisher.test(matrix(
      c(False, True, total_false - False, total_true - True),
      nrow = 2
    ))$p.value
  ) %>%
  ungroup()

# Display results with species names and p-values
species_p_values
```


```{r}
# Calculate average abundance for each species and filter top 10
top_n_species <- rel_abun_long %>%
  group_by(species) %>%
  summarize(mean_abundance = mean(abundance, na.rm = TRUE)) %>%
  arrange(desc(mean_abundance)) %>%
  slice_head(n = 9) # Number of species analyzed

print(top_n_species)

top_n_species_names <- top_n_species %>%
  pull(species)

# Filter for only top average species and plot
top_n_species_data <- rel_abun_long %>%
  filter(species %in% top_n_species_names)

ggplot(top_n_species_data, aes(x = species, y = abundance, fill=species)) +
  geom_boxplot(outlier.alpha = 0.1) +
  scale_y_log10() +
  stat_summary(fun.y=mean, shape=18, size=1, color="red") +
  labs(title = "Top Species by Average Relative Abundance",
       x = "Top Species",
       y = "Relative Abundance (log scale)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(top_n_species_data, aes(x = species, y = abundance, fill = species)) +
  geom_violin(alpha = 0.7) +
  scale_y_log10() +
  stat_summary(fun.y=mean, shape=18, size=1, color="red") +
  labs(title = "Density of Relative Abundance per Species",
       x = "Species",
       y = "Relative Abundance (log scale)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Plot density plots for each of the top 5 species with fixed scales
ggplot(top_n_species_data, aes(x = abundance)) +
  geom_density(fill = "skyblue", alpha = 0.7) +
  scale_x_log10() +  # Log scale for abundance comparison
  labs(title = "Density Plots of Most Abundant Species",
       x = "Relative Abundance (log scale)",
       y = "Frequency") +
  facet_wrap(~ species, scales = "fixed") +  # Fixed scales for comparison
  theme_minimal()
```
